# Data Wrangling {#sec-wrangle}

```{r}
#| label: setup-wrangle
#| include: false
library(tidyverse)  # data wrangling 
library(readxl)    # read excel files
library(janitor)    # data table cleaning
```

## Intended Learning Outcomes {#sec-ilo-wrangle .unnumbered}

- [ ] Be able to select and filter data for relevance
- [ ] Be able to create new columns and edit existing ones
- [ ] Be able to handle missing data

## Functions used {#functions-wrangle}

* built-in (you can always use these without loading any packages)
    * base:: `round()`, `mean()`, `min()`, `max()`, ``sum()`, `as.character()`, `as.numeric()`
    * stats:: `median()`
    * utils:: `head()`, `packageVersion()`
* tidyverse (you can use all these with `library(tidyverse)`)
    * readr:: `read_csv()`
    * dplyr:: `count()`, filter()`, `arrange()`, `mutate()`, `summarise()`, `group_by()`, `ungroup()`, `case_when()`, `na_if()`, `across()`
    * tidyselect:: `starts_with()`, `ends_with()`, `contains()`, `matches()`, `num_range()`, `all_of()`, `any_of()`, `everything()`, `where()`
* other (you need to load each package to use these)
    * readxl:: `read_xlsx()`
    * janitor:: `clean_names()`, `round_half_up()`

## Set-up {#sec-setup-wrangle}

::: {.callout-tip .activity}

1. Open your `ADS-2026` project 
1. Download the [Data transformation cheat sheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html)
1. Download a data file into the "data" folder: 
    * <a href="https://psyteachr.github.io/ads-v2/data/SDG-11.2.1.xlsx" download>SDG-11.2.1.xlsx</a>
1. Create a new quarto file called `09-wrangle.qmd`
1. Update the YAML header 
1. Replace the setup chunk with the one below

````
```{{r}}
#| label: setup
#| include: false

library(tidyverse) # for data wrangling
library(readxl)    # read excel files
library(janitor)   # clean up data tables
```
````

:::

::: {.callout-warning}
At this point in ADS it's very likely you will have the most recent version of the <pkg>tidyverse</pkg> packages but just in case, be aware that the function `case_when()` that we will use in this chapter was updated in to introduce new arguments and ways of handling `NA`. The code we provide in this chapter will only work if you have v1.1.0 or above of <pkg>dplyr</pkg> installed. To check run `packageVersion("dplyr")` and if it's below 1.1.0, install it again to update.
:::

## Wrangling functions

Data wrangling refers to the process of cleaning, transforming, and restructuring your data to get it into the format you need for analysis and it's something you will spend an awful lot of time doing. Most `r glossary("data wrangling")` involves the reshaping functions you learned in @sec-tidy and six functions from the <pkg>dplyr</pkg> package that is loaded as part of the <pkg>tidyverse</pkg>: `select`, `filter`, `arrange`, `mutate`, `summarise`, and `group_by`. We covered the last two in detail in @sec-summary, and the others have already come up several times, so we'll focus on more advanced data wrangling.

It's worth highlighting that in this chapter we're going to cover these common functions and common uses of said functions. However, <pkg>dplyr</pkg> (and packages beyond it) has a huge number of additional wrangling functions and each function has many different arguments. Essentially, if you think you should be able to wrangle your data in a particular way that we haven't explicitly shown you, you almost certainly can, it might just take a bit of Googling to find out how. 

```{r}
#| echo: false
#| eval: false

```

We'll use data from the United Nations [SDG Indicators Database](https://unstats.un.org/sdgs/dataportal/database) on sustainable development goal 11.2.1: Proportion of population that has convenient access to public transport, by sex, age and persons with disabilities. You can access [metadata about this data](https://unstats.un.org/sdgs/metadata/files/Metadata-11-02-01.pdf) to find out more.

```{r}
transport <- readxl::read_xlsx("data/SDG-11.2.1.xlsx")  |>
  janitor::clean_names()
```


### Explore

After you load the data, use `glimpse(budget)` or `View(transport)` to get familiar with the data.

```{r}
#| echo: false
glimpse(transport)
```

It is useful to use the `count()` function to explore data, especially where there are too many rows or columns to view on one screen.


```{r}
count(transport, age, sex)
```

It looks like most of the data is for all ages and both sexes, so if we want to look at gender disparities in access to public transport, we will need to narrow down the data.

::: {.callout-tip .activity}
Use `count()` (or other functions) to figure out how many years and how many countries are represented in the data.

```{r}
#| webex.hide: true

# two ways to get the number of unique values in a column
n_countries <- transport |> count(geo_area_name) |> nrow()
n_years <- transport$time_period |> unique() |> length()
```

:::

There are also several columns that seem to have no data, but sometimes there are important values later in the dataset. We can use count to confirm that these columns are entirely NAs.

```{r}
count(transport, goal)
count(transport, target)
count(transport, indicator)
count(transport, series_code)
count(transport, series_description)
count(transport, reporting_type)
count(transport, units)
```


:::{.callout-tip .activity}
Other columns seems to have the exact same value for all rows. Double check this.


```{r}
#| webex.hide: true
#| eval: false
count(transport, time_coverage)
count(transport, upper_bound) 
count(transport, lower_bound) 
count(transport, base_period)
count(transport, geo_info_url) 
count(transport, custom_breakdown)
```

:::


### Select

You can select a subset of the columns (variables) in a table to make it easier to view or to prepare a table for display. You can also select columns in a new order.

#### By name or index

You can select columns by name or number (which is sometimes referred to as the column index). Selecting by number can be useful when the column names are long or complicated.

```{r}
# select single column by name
value <- transport |> select(value) 

# select single column by number
value <- transport |> select(9) 
```

You can select each column individually, separated by commas (e.g., `age, sex`) but you can also select all columns from one to another by separating them with a colon (e.g., `goal:value`).

The colon notation can be much faster because you don't need to type out each individual variable name, but make sure that you know what order your columns are in and always check the output to make sure you have selected what you intended.

```{r select-name}
# select columns individually
country_year <- transport |> 
  select(geo_area_name, time_period, value)

# select columns with colon
country_year <- transport |> 
  select(geo_area_name:value)
```

You can rename columns at the same time as selecting them by setting `new_name = old_col`. 

```{r select-newname}
country_year <- transport |>
  select(country = geo_area_name, 
         year = time_period, 
         value)

head(country_year, 2)
```

#### Un-selecting columns

You can select columns either by telling R which ones you want to keep as in the previous examples, or by specifying which ones you want to exclude by using a minus symbol to un-select columns. You can also use the colon notation to de-select columns, but to do so you need to put parentheses around the span first, e.g., `-(time_coverage:base_period)`, not `-time_coverage:base_period`.

```{r select-unselect}
# de-select individual columns
transport_clean <- transport |>
  select(-time_coverage, 
         -upper_bound, 
         -lower_bound, 
         -base_period, 
         -geo_info_url, 
         -custom_breakdown)

# de-select a range of columns
transport_clean <- transport |>
  select(-(time_coverage:base_period), 
         -geo_info_url, 
         -custom_breakdown)
```

#### Select helpers

Finally, you can select columns based on criteria about the column names.

| function | definition |
|----------|------------|
| `everything()` | all columns |
| `starts_with()` | select columns that start with a character string|
| `ends_with()` | select columns that end with a character string |
| `contains()` | select columns that contain a character string |
| `matches()` | select columns with a name that matches a regular expression pattern |
| `num_range()` | select columns with a name that matches the pattern `prefix` |
| `where()` | select columns where the function returns TRUE |
| `all_of()` | select all of the columns in a vector |
| `any_of()` | select any of the columns in a vector that exist in the table |

Below, we make a vector of column names for columns that have only one value.

```{r}
one_val_cols <- c(
  "goal",
  "target",
  "indicator",
  "series_code",
  "series_description",
  "time_coverage",
  "upper_bound",
  "lower_bound",
  "base_period",
  "geo_info_url",
  "custom_breakdown",
  "reporting_type",
  "units"
)
```

Now we can use this vector to unselect these columns.

```{r}
transport_clean <- transport |>
  select(-all_of(one_val_cols))
```

::: {.callout-tip collapse=true}

There is a way to do this without having to type in all of the column names, but rather use functions to determine which columns have only one value and get a
vector of their names. This is probaly overkill for small tables, but invaluable if you have a table with hundreds of columns, or you need to do this to many different tables.

```{r}
n_values <- transport |> 
  summarise(across(                    # <1>
    .cols = everything(),              # <2>
    .fns = \(col) unique(col) |> length() # <3>
  ))
one_val <- n_values == 1               # <4>
one_val_cols <- one_val[, one_val] |>  # <5>
  names()                              # <6>
```

1. Use `across()` to apply a function to several columns
2. Use `everything()` to select all columns
3. Create an anonymous function (`\(col)`) that makes a vector of the unique values for each column and returns its length
4. Check which values are equal to 1
5 Choose the columns where this is TRUE
6. Get the names of these columns

:::

```{r select-helpers, echo = FALSE}
x1 <- transport |> select(starts_with("geo_"))
x2 <- transport |> select(ends_with("_bound"))
x3 <- transport |> select(contains("_area_"))
x4 <- transport |> select(where(is.numeric))

a <- list(names(x1), names(x2), names(x3), names(x4)) |>
  sapply(paste, collapse = ", ")

a1 <- setNames(a, c("answer", "x", "x", "x")) |> longmcq()
a2 <- setNames(a, c("x", "answer", "x", "x")) |> longmcq()
a3 <- setNames(a, c("x", "x", "answer", "x")) |> longmcq()
a4 <- setNames(a, c("x", "x", "x", "answer")) |> longmcq()
```


::: {.callout-tip .activity}
What are the resulting columns for these four examples?

* `transport |> select(contains("_area_"))`
    `r a3`
* `transport |> select(where(is.numeric))`
    `r a4`
* `transport |> select(starts_with("geo_"))`
    `r a1` 
* `transport |> select(ends_with("_bound"))`
    `r a2`
:::


### Rename and relocate

Whist you can rename and reorder columns using `select()`, sometimes it is easier to use the `rename()` and `relocate()` functions.

```{r}
transport_clean <- transport |>
  select(-all_of(one_val_cols)) |>
  rename(year = time_period,
         country = geo_area_name) |>
  relocate(age, sex, .after = value) |>
  relocate(geo_area_code, .before = cities)
```

```{r}
#| echo: false
head(transport_clean)
```



### Filter

Whilst `select()` chooses the columns you want to retain, `filter()` chooses the rows to retain by matching row or column criteria.

You can filter by a single criterion. This criterion can be rows where a certain column's value matches a character value (e.g., "North") or a number (e.g., 9003). It can also be the result of a logical equation (e.g., keep all rows with a specific column value larger than a certain value). The criterion is checked for each row, and if the result is FALSE, the row is removed. You can reverse equations by specifying `!=` where `!` means "not".

```{r}
#| label: filter-examaple
#| eval: false

# select all rows where country is Togo
transport_clean |> filter(country == "Togo")

# select all rows where country is NOT Togo
transport_clean |> filter(country != "Togo")

# select all rows where value was more than 99
transport_clean |> filter(value > 99)

# select rows where sex is FEMALE or MALE
transport_clean |> filter(sex %in%  c("FEMALE", "MALE"))
```

::: {.callout-warning}
Remember to use `==` and not `=` to check if two things are equivalent. A single `=` assigns the right-hand value to the left-hand variable (much like the `<-` operator).
:::


```{r filter-mcq, echo = FALSE}
class <- tibble(id = 1:4, 
               grade = c("A", "A", "C", "B"),
               score = c(95, 91, 76, 84)
               )
x1 <- which(class$grade == "A")
x2 <- which(class$score == 91)
x3 <- which(class$score < 80)
x4 <- which(class$grade != "A")

a <- list(x1, x2, x3, x4) |>
  sapply(paste, collapse = ", ")

a1 <- setNames(a, c("answer", "x", "x", "x")) |> mcq()
a2 <- setNames(a, c("x", "answer", "x", "x")) |> mcq()
a3 <- setNames(a, c("x", "x", "answer", "x")) |> mcq()
a4 <- setNames(a, c("x", "x", "x", "answer")) |> mcq()
```


::: {.callout-tip .activity}
Which IDs are kept from the table below?

```{r, echo = FALSE}
class
```

* `filter(demo, score < 80)`
    `r a3`
* `filter(demo, grade == "A")`
    `r a1`
* `filter(demo, grade != "A")`
    `r a4`
* `filter(demo, score == 91)`
    `r a2`
    

:::

You can also select on multiple criteria by separating them by commas (rows will be kept if they match *all* criteria). Additionally, you can use `&` ("and") and `|` ("or") to create complex criteria.

```{r}
#| label: filter-symbols
#| eval: false

# only data from Togo in 2020
togo_2020 <- transport_clean |> 
  filter(
    country == "Togo",
    year == 2020
  )

# the same as above, using & instead of a comma
togo_2020 <- transport_clean |> 
  filter(
    country == "Togo" &
    year == 2020
  )

# values under 10 or over 90
extreme_values <- transport_clean |>  
  filter( value < 10 | value > 90 )
```

If you want the filter to retain multiple specific values in the same variable, the "match operator" (`%in%`) should be used rather than `|` (or). The `!` can also be used in combination here, but it is placed before the variable name.

```{r}
#| label: match-operator
#| eval: false
islands <- c("Solomon Islands", "Northern Mariana Islands")

only_islands <- transport_clean |>
  filter(country %in% islands)

no_islands <- transport_clean |>
  filter(!country %in% islands)
```

`r glossary("Operator")`	|Name   |is TRUE if and only if
-----------|----------------------|---------------------------------
`A < B`    |less than 	          |A is less than B
`A <= B`   |less than or equal    |A is less than or equal to B
`A > B`    |greater than 	        |A is greater than B
`A >= B`   |greater than or equal |A is greater than or equal to B
`A == B`   |equivalence 	        |A exactly equals B
`A != B`   |not equal 	          |A does not exactly equal B
`A %in% B` |in 	                  |A is an element of vector B

Finally, you can also pass many other functions to filter. For example, the package <pkg>stringr</pkg> that is loaded as part of the <pkg>tidyverse</pkg> contains many different functions for working with `r glossary("string", "strings")` (character data). For example, you you use `str_detect()` to only retain rows where the country contains the word "Island".

```{r}
transport_clean |>
  filter(str_detect(country, "Island"))
```

Note that `str_detect()` is case sensitive so it would not return values of "island" or "ISLAND". You can use the function `tolower()` or `toupper()` to convert a string to lowercase or uppercase before you search for substring if you need case-insensitive matching.

::: {.callout-warning}
`filter()` is incredibly powerful and can allow you to select very specific subsets of data. But, it is also quite dangerous because when you start combining multiple criteria and operators, it's very easy to accidentally specify something slightly different than what you intended. **Always check your output**. If you have a small dataset, then you can eyeball it to see if it looks right. With a larger dataset, you may wish to compute summary statistics or count the number of groups/observations in each variable to verify your filter is correct. There is no level of expertise in coding that can substitute knowing and checking your data. 
:::

Let's narrow down our dataset to just the rows that report values aggregated by age, but separate for sex.

```{r}
transport_filtered <- transport_clean |> 
  filter(sex %in% c("FEMALE", "MALE"),
         age == "ALLAGE")
```

::: {.try}
What check would you do to make sure the filtering above worked as intended?

```{r}
#| webex.hide: true
# this should produce only two row, 
# with MALE and FEMALE as sex 
# and ALLAGE as age
count(transport_filtered, sex, age)
```

:::

### Arrange {#sec-arrange}

You can sort your dataset using `arrange()`. You will find yourself needing to sort data in R much less than you do in Excel, since you don't need to have rows next to each other in order to, for example, calculate group means. But `arrange()` can be useful when preparing data for display in tables. `arrange()` works on character data where it will sort alphabetically, as well as numeric data where the default is ascending order (smallest to largest). Reverse the order using `desc()`.

```{r arrange}
# arranging the table 
# first by country in alphabetical order
# then by year in reverse order
transport_arranged <- transport_filtered |>
  arrange(country, desc(year))
```


::: {.callout-tip .activity}
If you want to sort character data/categories in a specific order, turn the column into a `r glossary("factor")` and set the `levels` in the desired order.

```{r}
transport_factor <- transport_filtered |>
  mutate(sex = factor(sex, levels = c("MALE", "FEMALE")))
```


```{r}
transport_factor |>
  filter(year %in% 2001:2003) |>
  select(country, year, sex) |>
  arrange(year, sex)
```

:::

### Mutate

The function `mutate()` allows you to add new columns or change existing ones by overwriting them by using the syntax `new_column = operation`.  You can add more than one column in the same mutate function by separating the columns with a comma. Once you make a new column, you can use it in further column definitions. For example, the following code creates a new column called `urban`, which is true if the string "Urban average" is detected in the `foot_note` column, and fixes the data type of `value` by replacing it with the numeric version.

```{r mutate}
transport_data <- transport_factor |>
  mutate(
    urban = str_detect(foot_note, "Urban average"),
    value = as.numeric(value)
  )
```

`mutate()` can also be used in conjunction with other functions and Boolean operators. For example, we can add another column to `transport_data` that states whether the value is above the median value. Just like when we used `r glossary("boolean expression", "Boolean expressions")` with filter, it will evaluate the equation and return TRUE or FALSE depending on whether the observation meets the criteria.

```{r}
transport_data <- transport_factor |>
  mutate(
    urban = str_detect(foot_note, "Urban average"),
    value = as.numeric(value),
    above_median = value > median(value)
  )
```

::: {.callout-warning}
You can overwrite a column by giving a new column the same name as the old column (see `region` or `product`) above. Make sure that you mean to do this and that you aren't trying to use the old column value after you redefine it.
:::

You can also use `case_when()` to specify what values to return, rather than defaulting to TRUE or FALSE:

```{r}
transport_data <- transport_factor |>
  mutate(
    urban = str_detect(foot_note, "Urban average"),
    value = as.numeric(value),
    category = case_when(
      value < 70 ~ "low",
      value < 85 ~ "average",
      value >=85 ~ "high",
      .default = NA
    )
  )

```

The `case_when()` function allows allows you to set multiple criteria, although we're only using one non-default criterion here. It can be very useful, but takes a little practice.

You can also use `case_when()` to combine different criteria:

```{r}
# use different cutoffs for urban and non-urban areas
transport_data <- transport_factor |>
  mutate(
    urban = str_detect(foot_note, "Urban average"),
    value = as.numeric(value),
    category = case_when(
      urban & value < 75 ~ "low",
      urban & value < 85 ~ "average",
      urban & value >= 85 ~ "high",
      !urban & value < 20 ~ "low",
      !urban & value < 60 ~ "average",
      !urban & value >= 60 ~ "high",
      .default = NA
    )
  )

```

::: {.callout-tip}
Be mindful that `.default` uses `=` whilst the others use `~`. Emily has lost quite a lot of her time and sanity to not realising this. 
:::



Just like `filter()`, `mutate()` is incredibly powerful and the scope of what you can create is far beyond what we can cover in this book. 

### Summarise {#sec-dplyr-summarise}

You were introduced to the `summarise()` function in @sec-summary-summarise. This applies summary functions to an entire table (or groups, as you'll see in the next section).

Let's say we want to 

```{r}
transport_data |>
  summarise(mean = mean(value),
            min = min(value), 
            max = max(value))
```



### Group By {#sec-dplyr-groupby}

You were introduced to the `group_by()` function in @sec-grouping. For example, you can break down the summary statistics above by sex

```{r}
transport_data |>
  group_by(sex) |>
  summarise(
    mean = mean(value),
    min = min(value), 
    max = max(value)
  ) |>
  ungroup()
```

Newer code tends to use this format, however, if you just need to group the data for one function, you can use `.by`.

```{r}
transport_data |>
  summarise(
    .by = sex,
    mean = mean(value),
    min = min(value), 
    max = max(value)
  )
```

::: {.callout-tip}
Functions that let you create your own argument names, like `mutate()` and `summarise`, usually have all their other arguments start with `.` so you can create new columns with names like "data" or "by". See what happens if you accidentally use `by` instead of `.by`.

```{r}
#| warning: false

transport_data |>
  summarise(
    by = sex,
    mean = mean(value),
    min = min(value), 
    max = max(value)
  ) |> head()
```

:::


::: {.callout-tip .activity}
How would you find out the maximum value for each country, separately by sex?

```{r, results='asis', echo = FALSE}
c(x      = "<pre>transport_data |>\n  summarise(.by = value, country, max = max(sex)</pre>",
  answer = "<pre>transport_data |>\n  summarise(.by = c(sex, country), max = max(value)</pre>",
  x      = "<pre>transport_data |>\n  summarise(by = c(value, country), max = max(sex)</pre>",
  x      = "<pre>transport_data |>\n  summarise(.by = sex, country, max = max(value)</pre>") |>
  sample() |> webexercises::longmcq() |> cat()
```

:::

You can also use `group_by()` or `by` in combination with other functions. For example, `slice_max()` returns the top N rows, ordered by a specific variable.

```{r}
# return top 2 values 
transport_data |>
  slice_max(n = 2, order_by = value)
```

But this can be combined with `group_by()` to return the top sales for each region:

```{r}
# return top 2 values per sex
transport_data |>
  slice_max(n = 2, order_by = value, by = sex)
```

## Complications

### Rounding

Let's say we want to round all the values to the nearest percent and add a % sign.

```{r}
transport_data |>
  mutate(pcnt = round(value) |> paste0("%")) |>
  filter(str_detect(value, "\\.5")) |>
  select(value, pcnt)
```

If you look at the values above, you can see that the first three round up fro .5, and the last rounds down. What's going on!?

This may seem like a mistake, but R rounds .5 to the nearest even number, rather than always up, like you were probably taught in school. This prevents overestimation biases, since x.5 is *exactly* halfway between x and x+1, so there is no reason it should always round up.

```{r}
round(0.5)
round(1.5)
```

However, this might throw a monkey wrench into your own systems. For example, our school policy is to round up for course marks at x.5. The easiest solution is to use the `round_half_up()` function from the package <pkg>janitor</pkg>.

This should work as you'd expect.

```{r}
round_half_up(0.5)
round_half_up(1.5)
```

### Missing values {#sec-missing-values}

If you have control over your data, it is always best to keep missing values as empty cells rather than denoting missingness with a word or implausible number. If you used "missing" rather than leaving the cell empty, the entire variable would be read as character data, which means you wouldn't be able to perform mathematical operations like calculating the mean. If you use an implausible number (0 or 999 are common), then you risk these values being included in any calculations as real numbers.

However, we often don't have control over how the data come to us, so let's run through how to fix this.

#### Bad missing values

What if the South region hadn't returned their expenses (entered as 0) and the North region hadn't returned their sales data for 2020 yet, so someone entered it as "missing"?  

First, we're going to recode the data to add in the missing values. We can use `case_when()` to set the `value` to "missing" if the year is 2000 and country is "Uzbekistan", otherwise use the original value. You can't combine number and character data types in the same column, so you need to convert the default value into a character (thi is your first sign that this isn't a good idea).

```{r}
missing_bad <- transport_data |>
  mutate(value = case_when(
    # set to 0 when year is 2000 and country is Uzbekistan
    year == 2000 & country == "Uzbekistan" ~ "missing", 
    # otherwise, set to the existing value
    .default = as.character(value)
  ))
```


Now, if you try to compute the mean value, you will get an error message and the result will be `NA`.

```{r}
#| error: true
# try to compute mean value
missing_bad |>
  summarise(mean = mean(value))
```

#### Convert missing values to NA

To set the missing values to NA, we can use the handy function `na_if()`. We'll also need to transform `value` back to numeric.

```{r}
missing_data <- missing_bad |>
  mutate(
    # if value = "missing" set as NA
    value = na_if(value, "missing"),
    # convert to numeric
    value = as.numeric(value)
  )
```

Now, if we try to calculate the mean value for each country, we get missing values for any summary value that contains an NA.

```{r}
missing_data |>
  summarise(
    mean = mean(value),
    .by= country
  )
```

#### Ignore missing values

This is because `NA` basically means "I don't know", and the sum of 100 and "I don't know" is "I don't know", not 100. However, when you're calculating means, you often want to just ignore missing values. Set `na.rm = TRUE` in the summary function to remove missing values before calculating.

```{r}
missing_data |>
  summarise(
    mean = mean(value, na.rm = TRUE),
    .by= country
  )
```

#### Count missing values

If you want to find out how many missing or non-missing values there are in a column, use the `is.na()` function to get a `r glossary("logical")` vector of whether or not each value is missing, and use `sum()` to count how many values are TRUE or `mean()` to calculate the proportion of TRUE values.

```{r}
missing_data |>
  summarise(
    n_valid = sum(!is.na(value)),
    n_missing = sum(is.na(value)),
    prop_missing = mean(is.na(value)),
    .by= country
  )
```

#### Omit missing values

You may also want to remove rows that have missing values and only work from complete datasets. `drop_na()` will remove any row that has a missing observation. You can use `drop_na()` on the entire dataset which will remove any row that has *any* missing value, or you can specify to only remove rows that are missing a specific value.

```{r}
# remove any rows with any missing values
complete_data <- missing_data |>
  drop_na()

# remove any rows that are missing a value for sales
complete_values <- missing_data |>
  drop_na(value)

```

::: {.callout-warning}
If you use `drop_na()` without specifying columns, make absolutely sure that is what you want to do. There are often columns in a dataset that you don't care about, such as a "notes" column, that contain NAs, and these will be removed unless you specify which columns are the important ones.
:::

Missing data can be quite difficult to deal with depending on how it is represented. As always, no amount of coding expertise can make up for not understanding the structure and idiosyncrasies of your data. 

## Exercises {#sec-together-wrangle}

Let's try some exercises using a dataset with simulated survey data so that you can see how much more you're able to do with the data now.

* Save your current Markdown, close it, and open a new qmd named "survey_data_mad_skillz".
* In the set-up code chunk, load the tidyverse, then load the dataset from https://psyteachr.github.io/ads-v2/data/survey_data.csv into an object named `survey_data`.
* Use your method of choice to review the dataset and familiarise yourself with its structure.

```{r webex.hide = TRUE}
# from https://www.kaggle.com/kyanyoga/sample-sales-data
library(tidyverse)
survey_data <- read_csv("https://psyteachr.github.io/ads-v2/data/survey_data.csv")

```

### Creating new categories

Employees 1-5 were trained by Michael and employees 6-10 were trained by Dwight. 

* Create a new column named `trainer` that lists the trainer for each employee.
* Then, calculate the average satisfaction scores for employees trained by each trainer and visualise the satisfaction scores for each in whatever way you think best.

` r hide("Hint")`
To add the trainer column you can use `case_when()` and specify multiple criteria (e.g., if the employee is 1-5, Michael, if the employee is 6-10 Dwight)
` r unhide()`

```{r webex.hide = TRUE}
# case_when() method
survey_data <- survey_data |>
  mutate(trainer = case_when(
    employee_id %in% c("E01", "E02", "E03", "E04", "E05") ~ "Michael",
    employee_id %in% c("E06", "E07", "E08", "E09", "E10") ~ "Dwight")
  )

# mean satisfaction scores
survey_data |>
  summarise(.by = trainer,
            mean_satisfaction = mean(satisfaction))

# possible visualisation 

ggplot(survey_data, aes(x = satisfaction, fill = trainer)) +
  geom_histogram(binwidth = 1, show.legend = FALSE, colour = "black") +
  facet_wrap(~trainer) +
  labs(title = "Satisfaction scores by employee trainer")
```

### Filter by calculated score

First, calculate the average wait time and store this in an object named `mean_wait`. This should be a single value rather than a table.

`r hide("Hint")`
There are multiple ways to achieve this. You could create the table and then pull out the single value, or just calculate the single value.
`r unhide()`

```{r webex.hide = TRUE}

# method 1 - tidyverse
mean_wait <- survey_data |>
  summarise(mean_wait = mean(wait_time)) |>
  pull(mean_wait)

# method 2 - base R
mean_wait <- mean(survey_data$wait_time)

```

Now create a dataset named `long_wait` that just contains data from customers who waited more than the average wait time.

```{r webex.hide = TRUE}
long_wait <- survey_data |>
  filter(wait_time > mean_wait)
```

Create a visualisation that shows how many customers waited more than the average wait time for each employee.

```{r webex.hide = TRUE}
long_wait |>
  ggplot(aes(x = employee_id)) +
  geom_bar()
```

### Multiple critera

Now, add a column to `survey_data` named `follow_up` that flags whether a customer should be followed up with a courtesy phone call. Your company is short-staffed so only customers that meet all three of the following criteria should be followed-up:

* Their wait time should be above the average **for all calls**
* Their call time should be above the average **for their category**
* Their satisfaction should be less than three 3. 

This is quite complicated and there are multiple ways to achieve the desired outcome. Some approaches may need other functions that were covered in previous chapters and you may need to create intermediate objects.

Call the final object `follow_data` and keep only the customer ID, employee ID, trainer, and follow up columns. 

```{r webex.hide = TRUE}

# this is one possible solution, there are many other valid approaches 

# calculate mean wait time across all calls
mean_wait <- mean(survey_data$wait_time)

# calculate mean call time for each category
follow_data <- survey_data |>
  summarise(.by = issue_category,
            mean_call = mean(call_time)) |>
#then join it to the survey data  
  left_join(survey_data, by = "issue_category") |>
# then add on the column
  mutate(follow_up = case_when(wait_time > mean_wait & 
                               call_time > mean_call & 
                               satisfaction < 3 ~ "yes",
                               .default = "no")) |>
  select(caller_id, employee_id, trainer, follow_up)
```

For all of the above, write code that stores the answer as a single value, so that you could easily use it in inline coding.

How many customers need to be followed up:

* In total? `r fitb("120")`
* From calls by employee 06? `r fitb("16")`
* From calls by employees trained by Michael `r fitb("65")`
* From calls by employees trained by Dwight `r fitb("55")`

`r hide("Hint")`
`group_by |> count() |> filter() |> pull()
`r unhide()`

Which employee needs to make the largest number of follow-up courtesy calls? `r fitb("E02")`

`r hide("Hint")`
As above but add in an `ungroup()` and `slice_max()` along the way. 
`r unhide()`

```{r webex.hide = TRUE}
# in total
follow_data |>
  count(follow_up)|>
  filter(follow_up == "yes") |>
  pull(n)

# by employee 6
follow_data |>
  count(follow_up, employee_id) |>
  filter(employee_id == "E06",
         follow_up == "yes") |>
  pull(n)

# by michael
follow_data |>
  count(follow_up, trainer) |>
  filter(trainer == "Michael",
         follow_up == "yes") |>
  pull(n)

# by dwight
follow_data |>
  count(follow_up, trainer) |>
  filter(trainer == "Dwight",
         follow_up == "yes") |>
  pull(n)

# most follow-ups needed
follow_data |>
  count(follow_up, employee_id) |>
  filter(follow_up == "yes") |>
  slice_max(n = 1, order_by = n) |>
  pull(employee_id)

```

### Original insight

In preparation for the final summative assessment, explore the data to provide one original insight of your own.

### Share on Teams

Once you're done, share your rendered html file and the qmd file for the exercises on Teams in the Week 09 channel.


## Glossary {#sec-glossary-wrangle}

```{r}
#| echo: false
glossary_table()
```

## Further resources {#sec-resources-wrangle}

* [Data transformation cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf)
* [Chapter 18: Missing Data](https://r4ds.hadley.nz/missing-values) in *R for Data Science*
* [Chapter 3: Data Transformation ](https://r4ds.hadley.nz/data-transform) in *R for Data Science*
* [Chapter 25: Functions](https://r4ds.hadley.nz/functions) in *R for Data Science*
* [Introduction to stringr](https://stringr.tidyverse.org/articles/stringr.html)




